---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
# Title <- "Title of your paper goes here"
```

# STADEM

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/BiomarkABS/STADEM/master?urlpath=rstudio)

**ST**ate space **A**dult **D**am **E**scapement **M**odel

## Description

The primary goal of this project is to estimate total adult escapement of spring/summer Chinook salmon and steelhead that cross Lower Granite dam (LGD). In addition, to meet desired management and research objectives, total escapement has to include estimates of uncertainty and be parsed into weekly strata by three origin groups; wild, hatchery and hatchery no-clip. To reach this goal, we have developed the **ST**ate space **A**dult **D**am **E**scapement **M**odel (STADEM) model that incorporates fish ladder window counts, data from sampled fish at the LGD adult trap, and observations of previously PIT tagged fish at LGD adult detection sites.

This package contains functions to query and summarise the necessary data to fit a STADEM model, as well as write the JAGS model and run it using the jagsUI package in R. Some of the functions can be applied to other dams, and we are actively working to be able to run STADEM at locations other than Lower Granite Dam.

## Installation instructions

`STADEM` requires several packages that are available through the `tidyverse` package. You can install all the necessary packages by using:
  
```{r, eval = FALSE}
install.packages("tidyverse", "rjags", "jagsUI")
```

To install `STADEM` you can use Hadley Wickham's `devtools` package. To install and load the devtools package use:

```{r, eval = FALSE}
install.packages("devtools")
library(devtools)
```

NOTE: To use devtools, you may also have to download and install Rtools (although you shouldn't). The latest version on Rtools can be found at https://cran.r-project.org/bin/windows/Rtools/
  
You can download the compendium as a zip from from this URL: <https://github.com/BiomarkABS/STADEM/archive/master.zip>

Or you can install this compendium as an R package, STADEM, from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
remotes::install_github("BiomarkABS/STADEM", build_vignettes = TRUE)
```

STADEM requires the JAGS software (**J**ust **A**nother **G**ibbs **S**ampler). This can be downloaded here:
  
[https://sourceforge.net/projects/mcmc-jags/files/](https://sourceforge.net/projects/mcmc-jags/files/)
  
Please download version >= 4.0.0

Further instructions on how to use STADEM can be found in the package vignette, accessed by typing `browseVignettes(package = 'STADEM')`

## Authors

STADEM is a collaborative project, with the primary contributors being:
  
* Kevin See (Biomark, Inc.)
* Ryan N. Kinzer (Nez Perce Tribe)
* Rick Orme (Nez Perce Tribe)
* Mike Ackerman (Biomark, Inc.)


### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

